Global spirals in self-gravitating dead zones in protoplanetary discs

### 2014-10-23 paper/paper.tex
minor updates

\title[One-arm spirals in self-gravitating discs]{One-arm spirals in
  self-gravitating dead zones in protoplanetary discs}

@@

\maketitle
\begin{abstract}
  Theoretical models of circumstellar or protoplanetary discs indicate
  the possible existence of `dead zones' --- regions in the disc
  where mass accretion is greatly reduced because of inefficient
  magneto-hydrodynamic turbulence that would otherwise enable 
  accretion. It has been suggested that mass built-up in a dead zone
  may lead to gravitational instability. However, what type of
  gravitational instability is relevant in this context has not been
  clarified. In this work, we use numerical simulations to explore
  gravitational instability in astrophysical discs in which only a 
  sub-portion of it is self-gravitating, taken as a toy model of a 
  protoplanetary disc containing a massive dead zone. We conduct
  numerical hydrodynamic 
  experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
  and PLUTO codes. In all cases we observe the development of a spiral
  mode with azimuthal wavenumber $m=1$ inside the dead
  zone. Surprisingly, the co-rotation radius of the spiral is well
  outside the dead zone, resulting in a nearly stationary one-arm spiral
  pattern in the dead zone. Our simulations demonstrate the 
  importance of global gravitational interaction between a 
  self-gravitating dead zone and the exterior, non-self-gravitating
  parts of the disc.  
\end{abstract}

>>>>

\title[Spirals in self-gravitating dead zones]{Global spirals in
  self-gravitating dead zones in protoplanetary discs}

@@

\maketitle
\begin{abstract}
  Theoretical models of circumstellar or protoplanetary discs indicate
  the possible existence of `dead zones' --- regions in the disc
  where mass accretion is greatly reduced because of inefficient
  magneto-hydrodynamic turbulence that would otherwise enable 
  accretion. It has been suggested that mass built-up in a dead zone
  may lead to gravitational instability. However, what type of
  gravitational instability is relevant in this context has not been
  clarified. In this work, we use numerical simulations to explore
  gravitational instability in astrophysical discs in which only a 
  sub-portion of it is self-gravitating, taken as a toy model of a 
  protoplanetary disc containing a finite-size massive dead zone. We conduct
  numerical hydrodynamic 
  experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
  and PLUTO codes. In all cases we observe the development of a spiral
  mode with azimuthal wavenumber $m=1$ inside the dead
  zone, with co-rotation radius well
  outside the dead zone, resulting in a nearly stationary one-arm spiral
  pattern in the dead zone. Our simulations demonstrate the 
  importance of global gravitational interaction between a 
  self-gravitating dead zone and the exterior, non-self-gravitating
  parts of the disc. This non-local effect, due to disc structure, 
  exists even when the total disc mass is below the threshold for
  global effects found in
  previous studies. 
\end{abstract}

### 2014-10-23 paper/results.tex
minor updates

\section{Preliminary experiments}
Here we describe one numerical simulation from each code to give a
qualitative picture of disc evolution. In the next section, we analyse
2D simulations in more detail and explore parameter space.

In these simulations we subject the disc to random perturbations in
cylindrical velocity,
\begin{align}\label{randpert}
  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude and 
\begin{align}
  T(R) =
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
          R_d}\right)^2\right]}, 
\end{align}
where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
(R_{d2}-R_{d1})/2$. 
\subsection{FARGO 2D}\label{fargo_fiducial}
We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
[0.4,10]R_0$. This gives a total disc mass $M_{d}=0.086M_*$, and the
mass within $R\in[R_{d1},R_{d2}]$ is $0.049M_*$. 
We use a resolution of $N_R\times N_\phi = 512\times 1024$ or
about $8$ grids per $H$, and adopt $\epsilon_g=10^{-4}H$ for the 
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.
Perturbations are set initially with $M=10$ and
$\delta\in[-10^{-3},10^{-3}]$ chosen randomly but does not depend
on $\phi$.

@@

Fig. \ref{fargo_2d_angmom} compares the disc angular momenta
evolution, most of which is contained in the $m=0$ and $m=1$ modes. 
The $m=1$ spiral has an associated negative angular
momentum. Its growth is correlated with an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 \sim - \Delta
J_1$. Note that FARGO does not explicitly conserve angular
momentum. In addition, there may be an angular momentum flux due to
self-gravity across disc boundaries since our domain size is finite. 
Nevertheless, we find the total angular momentum is conserved to 
$\Delta J = O(10^{-7}J)$, and is much smaller in magnitude than the change in the
angular momenta components, $|\Delta J_m|$. 
Fig. \ref{fargo_2d_angmom} then indicates a transfer of 
angular momentum from the $m=1$ spiral to the background disc.

\begin{figure}
  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
  0cm 0cm
  0cm]{figures/polarxy_dens100}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm 0cm]{figures/polarxy_dens125} 
  \caption{Fiducial FARGO 2D simulation (\S\ref{fargo_fiducial}),
    showing the evolution of the non-axisymmetric surface density
    $\Delta\Sigma$. Note that the disc extends to $R=10R_0$, but only
    the inner part is shown for clarity. \label{fargo_2d}} 
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the FARGO
    fiducial simulation (Fig. \ref{fargo_2d}). The perturbation
    relative to $t=0$ is shown in units of the initial total angular
    momentum $J_\mathrm{ref}$.\label{fargo_2d_angmom}} 
\end{figure}

\subsection{ZEUS-MP 3D}
The 3D disc has radial size 
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
spiral in the dead zone, which was found to dominate the 2D
simulation.

@@

\subsection{PLUTO 3D}

>>>>

\section{Preliminary experiments}
Here we describe one numerical simulation from each code to give a
qualitative picture of disc evolution. 
%In the next section, we analyse
%2D simulations in more detail and explore parameter space with
%improved resolution runs.

In these simulations we subject the disc to random perturbations in
cylindrical velocity,
\begin{align}\label{randpert}
  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude and 
\begin{align}
  T(R) =
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
          R_d}\right)^2\right]}, 
\end{align}
where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
(R_{d2}-R_{d1})/2$. 
\subsection{2D run}\label{fargo_fiducial}
We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
[0.4,10]R_0$. This gives a total disc mass $M_{d}=0.086M_*$, and the
mass within $R\in[R_{d1},R_{d2}]$ is $0.049M_*$. 
We use a resolution of $N_R\times N_\phi = 1024\times 2048$ or
about $16$ grids per $H$, and adopt $\epsilon_g=10^{-4}H$ for the 
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.
Perturbations are set initially with $M=10$ and
$\delta\in[-10^{-3},10^{-3}]$ chosen randomly but does not depend
on $\phi$.

@@

Fig. \ref{fargo_2d_angmom} compares the disc angular momenta
evolution, most of which is contained in the $m=0$ and $m=1$ modes. 
The $m=1$ spiral has an associated negative angular
momentum. Its growth is correlated with an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 \sim - \Delta
J_1$. Note that FARGO does not explicitly conserve angular
momentum. In addition, there may be an angular momentum flux due to
self-gravity across disc boundaries since our domain size is finite. 
Nevertheless, we find the total angular momentum is conserved to 
$|\Delta J/J|= O(10^{-6})$, and is much smaller in magnitude than the change in the
angular momenta components, $|\Delta J_{0,1}/J|> O(10^{-5})$. 
Fig. \ref{fargo_2d_angmom} then indicates a transfer of 
angular momentum from the $m=1$ spiral to the background disc.

\begin{figure}
  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
  0cm 0cm
  0cm]{figures/polarxy_dens100}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm 0cm]{figures/polarxy_dens125} 
  \caption{Preliminary FARGO 2D simulation (\S\ref{fargo_fiducial}),
    showing the evolution of the non-axisymmetric surface density
    $\Delta\Sigma$. Note that the disc extends to $R=10R_0$, but only
    the inner part is shown for clarity. \label{fargo_2d}} 
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the FARGO
     simulation show in Fig. \ref{fargo_2d}. The perturbation
    relative to $t=0$ is shown in units of the initial total angular
    momentum $J_\mathrm{ref}$.\label{fargo_2d_angmom}} 
\end{figure}

\subsection{3D runs}
The 3D disc has radial size 
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
spiral in the dead zone, which was found to dominate the 2D
simulation.

@@

Snapshots at the end of the simulations are shown in
Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again
find the dead zone eventually dominated by an $m=1$ spiral. Note that
the amplitude of the spiral is smaller than the 2D case above,
possibily due to lower resolution and/or 3D self-gravity being weaker
than that in 2D.

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.33]{figures/polarxy_dens015_zeus}
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.33]{figures/polarxy_dens015_zeus}
    }
  \end{center}
  \caption{Preliminary 3D simulations using the (a) ZEUS-MP and (b)
    PLUTO. The non-axisymmetric midplane density at the end of the run
    is shown. Here $\psi \equiv \pi/2 - \theta$ is the angular height
    from the midplane.\label{3d_prelim}}   
\end{figure}

Fig. \ref{3d_prelim_angmom} shows the angular momentum evolution in
the 3D runs. ZEUS-MP does not conserve angular momentum very well, but
the variation $|\Delta J/J|< O(10^{-5})$ is not significant compared
to the individual components $|\Delta J_{0,1}/J|\sim
6\times10^{-5}$. We found high-$m$ structures to develop near the
inner boundary, which may have contributed to this error. It
is clear, however, that the growth of negative angluar momentum in
$m=1$ is accompanied by an increase in the background angular
momentum.

\begin{figure}
  \includegraphics[scale=0.41,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
  \includegraphics[scale=0.41]{figures/nonaxi_evol_ang_zeus}
  \caption{Evolution of angular momentum components in the 3D 
    simulations show in Fig. \ref{3d_prelim}. The perturbation
    relative to $t=10P_0$ is shown in units of the initial total
    angular momentum $J_\mathrm{ref}$.\label{3d_prelim_angmom}} 
\end{figure}

\section{Global one-arm spirals in structured discs}

\subsection{Properties of the $m=1$ dead zone spiral}

%\subsection{Three-dimensional structure}

%\section{Additional simulations}
%survery
